Automated Residual Plot Assessment with the R Package autovi and Shiny App autovi.web

### 2024-11-19 autovi_paper.qmd
done with cutting paper

## Numerical Summary {#sec-autovi-usage}

@@

## Visual Summary {#sec-autovi-visual}

@@

## Design {#sec-autovi-web-design}

```{r fig-autovi-web, out.width = "100%", fig.cap = "Overview of the `autovi.web` graphical user interface (GUI). This default view may change based on user interactions. Region 1 is the sidebar menu, containing the residual assessment tab and the information tab. Region 2 is the data upload panel, where users can provide a CSV file and specify the type of data it contains. Region 3 includes dropdown menus for selecting the columns to be analyzed, a slider to control the number of bootstrapping samples, and a numeric input box for setting the simulation seed. Region 4 displays the initialization status and offers a button to start the analysis. Region 5 is empty in the default view but will be populated with results once the analysis is started. "}
knitr::include_graphics("figures/autovi_web.png")
```

While the R package `autovi` aims to provide tools that can be extended to broader visual inference applications, `autovi.web` is only focus on providing  a straightforward and clean user interface. An overview of the graphical user interface of `autovi.web` is provided in @fig-autovi-web. This is the default view of the web application, and there are five regions that user can mainly interact with. Region 1 of @fig-autovi-web is a sidebar menu which can switch between the analysis page and the information page. The analysis page is the focus of this section.

@@

The web application is designed to assess either a single residual plot or a lineup of residual plots. Therefore, it accepts only two types of CSV files: one with at least two columns representing the fitted values and residuals of a single residual plot, and another with at least three columns, where the additional column serves as the label or identifier for a lineup of multiple residual plots. For a single residual plot, 19 null plots are generated by simulating normal random draws from a distribution with the same variance as the original residual plot, and comparisons are made with the original residual plot. For a lineup, comparisons are made among the plots within the lineup. After uploading the CSV file, the user must select the correct format to ensure the web interface interprets the data correctly.

```{r fig-autovi-web-type, out.width = "100%", fig.cap = "The panels for selecting target columns and simulation settings are updated when a different CSV type is selected in the left panel. Compared to @fig-autovi-web, where the CSV type is a single residual plot, choosing a CSV type that includes a lineup of multiple residual plots adds a dropdown menu for specifying a column for the residual plot identifier. Additionally, an optional dropdown menu for specifying the true residual plot identifier will appear under the simulation settings."}
knitr::include_graphics("figures/autovi_web_type.png")
```

Region 3 of @fig-autovi-web is a panel for column selection and simulation settings. As shown in @fig-autovi-web, if the CSV type is set to a single residual plot, there will be two dropdown menus for specifying the columns for fitted values and residuals, respectively. The default variable names for these columns are `.fitted` and `.resid`. After uploading the CSV file, the content of these dropdown menus will be updated to reflect the existing columns in the dataset. As displayed in @fig-autovi-web-type, for the CSV type that is a lineup of multiple residual plots, an additional dropdown menu will appear for specifying the column of residual plot labels. The default variable name for this column is `.sample`. If this variable name does not exist in the dataset, the dropdown menu will remain empty, allowing the user to specify the correct column. The number of levels for each option in this dropdown menu will be displayed to help avoid the selection of a variable with too many levels, which could significantly slow down the application due to extensive computation.

@@

Once the play button is clicked, region 5 of @fig-autovi-web will be populated with panels displaying the assessment results. Generally, there will be four result panels, as shown in @fig-autovi-web-result and @fig-autovi-web-result2.

```{r fig-autovi-web-result, out.width = "100%", fig.cap = "The first two panels of results from the automated residual assessment are shown. The application provides four results panels in total, and these screenshots display the first two. In region 1, there is an interactive table detailing the VSS, with a summary of the analysis provided in the paragraph below. Region 2 displays a lineup of residual plots."}
knitr::include_graphics("figures/autovi_web_result.png")
```

```{r fig-autovi-web-result2, out.width = "100%", fig.cap = "The last two panels of results from the automated residual assessment are shown. The application provides four results panels in total, and these screenshots display the final two. Region 1 presents a density plot comparing the bootstrapped VSS with the null VSS. Region 2 includes an attention map of the true residual plot."}
knitr::include_graphics("figures/autovi_web_result2.png")
```

```{r fig-autovi-web-gradient-hide, out.width = "50%", fig.cap = "The attention map is hidden if the assessment indicates a $p$-value greater than 0.05. A button is available to toggle the display of the attention map."}
knitr::include_graphics("figures/autovi_web_gradient_hide.png")
```

Region 6 of @fig-autovi-web-result contains an interactive table created with the R package `DT` [@dt], which provides the VSS. This table includes four columns: `.sample`, `vss`, `rank`, and `null`. The `.sample` column shows the residual plot labels. For a CSV type that is a lineup, these labels are taken from an identifier column in the dataset specified by the user. In the case of the CSV type is a single residual plot, labels are automatically generated from 1 to 20, with the true residual plot receiving a randomly assigned label. The `vss` column displays the VSS for each residual plot, rounded to three decimal places. The `rank` column indicates the ranking of each residual plot based on VSS. The `null` column reveals whether the plot is a null plot. For the CSV type that is a single residual plot, only the true residual plot will have "false" in this column, while all other plots will be marked "true." For the CSV type that is a lineup, if the true residual plot identifier has not been provided, this column will show "NA" to represent missing values. If the identifier is provided by user, the column behaves as if the CSV type is a single residual plot.

@@

Region 9 of @fig-autovi-web-result2 displays an attention map for the true residual plot, generated by computing the gradient of the Keras model's output with respect to the greyscale input of the plot. The attention map helps to understand how the Keras model predicts VSS and which areas it is focusing on. We use a greyscale input because it is easier to generate a clear attention map in this format, and it usually conveys all the essential information, as most of the important details of the plot are drawn in black. If the $p$-value of the true residual plot is greater than 0.05, checking the attention map is not necessary. However, to provide users with the option to review it if they wish, a button will be available, as shown in @fig-autovi-web-gradient-hide. This button allows users to toggle the display of the attention map.

## Workflow {#sec-autovi-web-workflow}

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in each panel shown in @fig-autovi-web. There are two example datasets provided by the web application, as mentioned in @sec-autovi-web-design. The single residual plot example uses the `dino` dataset from the R package `datasauRus` [@datasaurus]. The lineup example uses residuals from a simulated regression model that has a non-linearity issue. We will walk through the lineup example to further demonstrate the workflow of the web application.

As shown in @fig-autovi-web-workflow-example1, to use the lineup example data, click the "Use Lineup Example" button. The data status will then update to show the number of rows and columns in the dataset, and the CSV type will automatically be selected to the correct option. Since the example dataset follows the variable naming conventions assumed by the web application, the columns for fitted values, residuals, and labels of residual plots are set automatically (@fig-autovi-web-workflow-example2). If the user is working with a custom dataset, these options must be set accordingly. Regardless of the dataset, the user must manually select the label for the true residual plot, as the web application allows assessments without this label. The next step is to click the play button to start the assessment.

Results are provided in multiple panels as displayed in @fig-autovi-web-workflow-example3, @fig-autovi-web-workflow-example4, @fig-autovi-web-workflow-example5 and @fig-autovi-web-workflow-example6. In @fig-autovi-web-workflow-example3, the first row of the table is the most crucial to check, as it provides the VSS and the rank of the true residual plot among the other plots. The summary text beneath the table provides the $p$-value, which can be used for quick decision-making. In @fig-autovi-web-workflow-example4, the lineup is for manual inspection, and the user should see if the true residual plot is visually distinguishable from the other plots, to confirm if the model violation is serious. The density plot in @fig-autovi-web-workflow-example5 offers a more robust result, allowing the user to compare the distribution of bootstrapped VSS with the distribution of null VSS. Finally, the grayscale attention map shown in @fig-autovi-web-workflow-example6 can be used to check if the target visual features, like the non-linearity present in the lineup example, are captured by the computer vision model, ensuring the quality of the assessment.

```{r fig-autovi-web-workflow-example1, out.width = "80%", fig.cap = 'To begin the workflow for `autovi` using the lineup example dataset, the user clicks the "Use Lineup Example" button to load the example dataset, during which the data status and CSV type will be automatically updated.'}
magick::image_read_pdf("figures/autovi_web_workflow_1.pdf")
```

```{r fig-autovi-web-workflow-example2, out.width = "80%", fig.cap = 'After clicking the button in @fig-autovi-web-workflow-example1, the target columns are selected automatically, though the user must manually select the label for the true residual plot, as the web application permits assessment without this label.'}
magick::image_read_pdf("figures/autovi_web_workflow_2.pdf")
```

```{r fig-autovi-web-workflow-example3, out.width = "55%", fig.cap = 'After finishing the required steps in @fig-autovi-web-workflow-example1 and @fig-autovi-web-workflow-example2, the user initiates the assessment of the lineup example data by clicking the run button.'}
magick::image_read_pdf("figures/autovi_web_workflow_3.pdf")
```

```{r fig-autovi-web-workflow-example4, out.width = "100%", fig.cap = 'The VSS of the true residual plot is displayed in the first row of the table, with a summary text beneath the table providing the $p$-value to aid in decision-making.'}
magick::image_read_pdf("figures/autovi_web_workflow_4.pdf")
```

```{r fig-autovi-web-workflow-example5, out.width = "100%", fig.cap = 'A lineup of residual plots allows for manual inspection.'}
magick::image_read_pdf("figures/autovi_web_workflow_5.pdf")
```

```{r fig-autovi-web-workflow-example6, out.width = "100%", fig.cap = 'The density plot helps verify if the bootstrapped distribution differs from the null distribution.'}
magick::image_read_pdf("figures/autovi_web_workflow_6.pdf")
```

```{r fig-autovi-web-workflow-example7, out.width = "100%", fig.cap = 'The attention map offers insights into whether the computer vision model has captured the intended visual features of the true residual plot.'}
magick::image_read_pdf("figures/autovi_web_workflow_7.pdf")
```

>>>>

## Usage {#sec-autovi-usage}

### Numerical Summary {#sec-autovi-numerical}

@@

### Visual Summary {#sec-autovi-visual}

@@

<!-- ## Design {#sec-autovi-web-design}-->

```{r fig-autovi-web, out.width = "100%", fig.cap = "Overview of the `autovi.web` graphical user interface (GUI). This default view may change based on user interactions. Region 1 is the sidebar menu, containing the residual assessment tab and the information tab. Region 2 is the data upload panel, where users can provide a CSV file and specify the type of data it contains. Region 3 includes dropdown menus for selecting the columns to be analyzed, a slider to control the number of bootstrapping samples, and a numeric input box for setting the simulation seed. Region 4 displays the initialization status and offers a button to start the analysis. Region 5 is empty in the default view but will be populated with results once the analysis is started. ", eval=FALSE}
knitr::include_graphics("figures/autovi_web.png")
```

<!--The focus of `autovi.web` is on providing a straightforward and clean user interface. An overview of the graphical user interface of `autovi.web` is provided in @fig-autovi-web. This is the default view of the web application, and there are five regions that user can mainly interact with. Region 1 of @fig-autovi-web is a sidebar menu which can switch between the analysis page and the information page. The analysis page is the focus of this section.

@@

The web application is designed to assess either a single residual plot or a lineup of residual plots. Therefore, it accepts only two types of CSV files: one with at least two columns representing the fitted values and residuals of a single residual plot, and another with at least three columns, where the additional column serves as the label or identifier for a lineup of multiple residual plots. For a single residual plot, 19 null plots are generated by simulating normal random draws from a distribution with the same variance as the original residual plot, and comparisons are made with the original residual plot. For a lineup, comparisons are made among the plots within the lineup. After uploading the CSV file, the user must select the correct format to ensure the web interface interprets the data correctly. -->

```{r fig-autovi-web-type, out.width = "100%", fig.cap = "The panels for selecting target columns and simulation settings are updated when a different CSV type is selected in the left panel. Compared to @fig-autovi-web, where the CSV type is a single residual plot, choosing a CSV type that includes a lineup of multiple residual plots adds a dropdown menu for specifying a column for the residual plot identifier. Additionally, an optional dropdown menu for specifying the true residual plot identifier will appear under the simulation settings.", eval=FALSE}
knitr::include_graphics("figures/autovi_web_type.png")
```

<!--Region 3 of @fig-autovi-web is a panel for column selection and simulation settings. As shown in @fig-autovi-web, if the CSV type is set to a single residual plot, there will be two dropdown menus for specifying the columns for fitted values and residuals, respectively. The default variable names for these columns are `.fitted` and `.resid`. After uploading the CSV file, the content of these dropdown menus will be updated to reflect the existing columns in the dataset. As displayed in @fig-autovi-web-type, for the CSV type that is a lineup of multiple residual plots, an additional dropdown menu will appear for specifying the column of residual plot labels. The default variable name for this column is `.sample`. If this variable name does not exist in the dataset, the dropdown menu will remain empty, allowing the user to specify the correct column. The number of levels for each option in this dropdown menu will be displayed to help avoid the selection of a variable with too many levels, which could significantly slow down the application due to extensive computation.

@@

Once the play button is clicked, region 5 of @fig-autovi-web will be populated with panels displaying the assessment results. Generally, there will be four result panels, as shown in @fig-autovi-web-result and @fig-autovi-web-result2. 
-->

```{r fig-autovi-web-result, out.width = "100%", fig.cap = "The first two panels of results from the automated residual assessment are shown. The application provides four results panels in total, and these screenshots display the first two. In region 1, there is an interactive table detailing the VSS, with a summary of the analysis provided in the paragraph below. Region 2 displays a lineup of residual plots.", eval=FALSE}
knitr::include_graphics("figures/autovi_web_result.png")
```

```{r fig-autovi-web-result2, out.width = "100%", fig.cap = "The last two panels of results from the automated residual assessment are shown. The application provides four results panels in total, and these screenshots display the final two. Region 1 presents a density plot comparing the bootstrapped VSS with the null VSS. Region 2 includes an attention map of the true residual plot.", eval=FALSE}
knitr::include_graphics("figures/autovi_web_result2.png")
```

```{r fig-autovi-web-gradient-hide, out.width = "50%", fig.cap = "The attention map is hidden if the assessment indicates a $p$-value greater than 0.05. A button is available to toggle the display of the attention map.", eval=FALSE}
knitr::include_graphics("figures/autovi_web_gradient_hide.png")
```

<!-- Region 6 of @fig-autovi-web-result contains an interactive table created with the R package `DT` [@dt], which provides the VSS. This table includes four columns: `.sample`, `vss`, `rank`, and `null`. The `.sample` column shows the residual plot labels. For a CSV type that is a lineup, these labels are taken from an identifier column in the dataset specified by the user. In the case of the CSV type is a single residual plot, labels are automatically generated from 1 to 20, with the true residual plot receiving a randomly assigned label. The `vss` column displays the VSS for each residual plot, rounded to three decimal places. The `rank` column indicates the ranking of each residual plot based on VSS. The `null` column reveals whether the plot is a null plot. For the CSV type that is a single residual plot, only the true residual plot will have "false" in this column, while all other plots will be marked "true." For the CSV type that is a lineup, if the true residual plot identifier has not been provided, this column will show "NA" to represent missing values. If the identifier is provided by user, the column behaves as if the CSV type is a single residual plot.

@@

Region 9 of @fig-autovi-web-result2 displays an attention map for the true residual plot, generated by computing the gradient of the Keras model's output with respect to the greyscale input of the plot. The attention map helps to understand how the Keras model predicts VSS and which areas it is focusing on. We use a greyscale input because it is easier to generate a clear attention map in this format, and it usually conveys all the essential information, as most of the important details of the plot are drawn in black. If the $p$-value of the true residual plot is greater than 0.05, checking the attention map is not necessary. However, to provide users with the option to review it if they wish, a button will be available, as shown in @fig-autovi-web-gradient-hide. This button allows users to toggle the display of the attention map.-->

## Usage {#sec-autovi-web-workflow}

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in each panel. There are two example datasets provided by the web application. The single residual plot example uses the `dino` dataset from the R package `datasauRus` [@datasaurus]. The lineup example uses residuals from a simulated regression model that has a non-linearity issue. We will walk through the lineup example to further demonstrate the workflow of the web application.

### Reading data and setting parameters

The user can select to upload data as either a single set of residuals or a pre-computed lineup of residuals and nulls. Here we illustrate use with lineup example data sets (@fig-autovi-web-setup).  To use the lineup example data, click the "Use Lineup Example" button. The data status will then update to show the number of rows and columns in the dataset, and the CSV type will automatically be selected to the correct option. Since the example dataset follows the variable naming conventions assumed by the web application, the columns for fitted values, residuals, and labels of residual plots are set automatically (middle image). If the user is working with a custom dataset, these options must be set accordingly. Regardless of the dataset, the user must manually select the label for the true residual plot, whenever a lineup file is provided, as the web application allows assessments without this label. The last step is to click the play button (right image) to start the assessment.

```{r fig-autovi-web-setup, out.width = "100%", fig.cap = 'To begin the workflow for `autovi` using the lineup example dataset, the user clicks the "Use Lineup Example" button (left) to load the example dataset, during which the data status and CSV type will be automatically updated. The target columns are selected automatically, though the user must manually select the label for the true residual plot (middle), as the web application permits assessment without this label. The user initiates the assessment of the lineup example data by clicking the run button (right).'}
magick::image_read_pdf("figures/autovi_web_setup.pdf")
```

```{r fig-autovi-web-workflow-example1, out.width = "80%", fig.cap = 'To begin the workflow for `autovi` using the lineup example dataset, the user clicks the "Use Lineup Example" button to load the example dataset, during which the data status and CSV type will be automatically updated.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_1.pdf")
```

```{r fig-autovi-web-workflow-example2, out.width = "80%", fig.cap = 'After clicking the button in @fig-autovi-web-workflow-example1, the target columns are selected automatically, though the user must manually select the label for the true residual plot, as the web application permits assessment without this label.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_2.pdf")
```

```{r fig-autovi-web-workflow-example3, out.width = "55%", fig.cap = 'After finishing the required steps in @fig-autovi-web-workflow-example1 and @fig-autovi-web-workflow-example2, the user initiates the assessment of the lineup example data by clicking the run button.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_3.pdf")
```

### Results provided

Results are provided in multiple panels.<!-- as displayed in @fig-autovi-web-workflow-example3, @fig-autovi-web-workflow-example4, @fig-autovi-web-workflow-example5 and @fig-autovi-web-workflow-example6. In @fig-autovi-web-workflow-example3--> The first row of the table [@fig-autovi-web-lineup] is the most crucial to check, as it provides the VSS and the rank of the true residual plot among the other plots. The summary text beneath the table provides the $p$-value, which can be used for quick decision-making. The lineup is for manual inspection, and the user should see if the true residual plot is visually distinguishable from the other plots, to confirm if the model violation is serious.

The density plot in @fig-autovi-web-distributions offers a more robust result, allowing the user to compare the distribution of bootstrapped VSS with the distribution of null VSS. Finally, the grayscale attention map (right image) can be used to check if the target visual features, like the non-linearity present in the lineup example, are captured by the computer vision model, ensuring the quality of the assessment.

```{r fig-autovi-web-lineup, out.width = "100%", fig.cap = 'Results for the lineup. The VSS of the true residual plot is displayed in the first row of the table of VSS values for all the null plots (left image), with a summary text beneath the table providing the $p$-value to aid in decision-making. A lineup of residual plots allows for manual inspection (right image). '}
magick::image_read_pdf("figures/autovi_web_lineup.pdf")
```

```{r fig-autovi-web-distributions, out.width = "100%", fig.cap = 'Summaries assessing the strength of the pattern and which elements of the plot contribute. The density plot helps verify if the bootstrapped distribution differs from the null distribution (left image). The attention map (right image) offers insights into whether the computer vision model has captured the intended visual features of the true residual plot.'}
magick::image_read_pdf("figures/autovi_web_distributions.pdf")
```

```{r fig-autovi-web-workflow-example4, out.width = "100%", fig.cap = 'The VSS of the true residual plot is displayed in the first row of the table, with a summary text beneath the table providing the $p$-value to aid in decision-making.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_4.pdf")
```

```{r fig-autovi-web-workflow-example5, out.width = "100%", fig.cap = 'A lineup of residual plots allows for manual inspection.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_5.pdf")
```

```{r fig-autovi-web-workflow-example6, out.width = "100%", fig.cap = 'The density plot helps verify if the bootstrapped distribution differs from the null distribution.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_6.pdf")
```

```{r fig-autovi-web-workflow-example7, out.width = "100%", fig.cap = 'The attention map offers insights into whether the computer vision model has captured the intended visual features of the true residual plot.', eval=FALSE}
magick::image_read_pdf("figures/autovi_web_workflow_7.pdf")
```
